---
title: "Common errors in R: An Empirical Investigation"
author: "Noam Ross"
date: "March 23, 2015"
output: 
  html_document:
    keep_md: true
---

`r library(knitr); opts_chunk$set(cache=TRUE)`

**tl; dr: Most errors in R are due to looking for something that isn't there.**

*(Comment in the issues [here](https://github.com/noamross/zero-dependency-problems/issues/19)!)*

In my project to [teach problem-solving skills in R](https://github.com/noamross/zero-dependency-problems), one area that has come
up is teaching [interpretation of common error messages](https://github.com/noamross/zero-dependency-problems/issues/7).  R has
notoriously difficult to decipher error messages.

Experienced users and instructors can come up with a list of common error
messages to teach, but how can one put oneself in a novice's shoes, and understand
what errors they have trouble understanding?

One approach is to look at forums such as [Stack Overflow](http://stackoverflow.com/),
where people ask questions about programming problems.
I spent some time clicking around looking for useful teaching examples.  It's actually
hard to find *novice* questions, though, because the site does its best
to direct people towards well-formed questions and answers. This is incredibly
useful when one is searching for answers, but it gives less of a view into what
kinds of problems people post onto the site.  The "long tail" of specialty
and poorly-formed questions is both harder to navigate and too big to read manually.

Thankfuly, Stack Overflow has a great API, and [David Robinson](http://varianceexplained.org/)
has written a great R package, [stackr](https://github.com/dgrtwo/stackr), to
access it.  Here I use stackr to find the most common types of error messages
R users post about on Stack Overflow.

First, get set up:

```{r setup}
#devtools::install_github("dgrtwo/stackr")
library(stackr)
library(stringi)
library(magrittr)

#Sys.setenv(STACK_EXCHANGE_KEY = "XXXXX") <-- Get a key or you'll hit the API
#                                             rate limits. See the stackr README

set.seed(0)
```


I searched the  `[r]` tag on SO for any questions with the word "Error" and got
[almost 20,000 results](http://stackoverflow.com/search?q=%5Br%5D+%22Error%22),
so I use stackr to repeat this query in R, making sure I repeat enough times
to get all the data, and return the full text of the questions with 
`filter="withbody"`.

```{r getresults}
results = stack_search(tagged = "r", body = "Error", filter="withbody",
                       pagesize=100, num_pages=200)
```

Extracting the error messages is a bit tricky. People copy and paste messages
in all sorts of ways, especially beginners.  After some trial and error, I came
up with a regex filter that grabs R error messages pretty well:

```{r matcherrors}
errors = stri_extract_all_regex(results$body,
  'Error[^:]*:.+?(?=(\\n\\n|In addition|Warning|</p>|</code>|</pre>|</blockquote>))',
   opts_regex=stri_opts_regex(mutliline=TRUE, dotall=TRUE))

errors = na.omit(unlist(errors))
errors = stri_replace_last_regex(errors, "\\n$", "")

length(errors)
length(unique(errors))
```

Now we have a list of over 10,000 errors, but how to derive some insight from
them?  The vast majority are unique because R uses some part of the command
in the error message itself.  I decided to use [n-grams](http://en.wikipedia.org/wiki/N-gram),
specifically, trigrams, to look at the most frequent word combinations in the error messages:


```{r trigrams}
trigrams = lapply(
  stri_split_regex(errors, "\\s+"),
  function(x) {
    len = length(x)
    if(length(x) < 3) return(NA)
    paste(x[1:(len-2)], x[2:(len-1)], x[3:len])
  }
)

trigram_freqs = table(unlist(trigrams)) %>%
  sort(decreasing = TRUE) %>%
  data.frame(trigram=names(.), freq = ., stringsAsFactors=FALSE)

rownames(trigram_freqs) = NULL
```

We can plot the distribution of trigrams.  Here are the top 100:

```{r plot, warning=FALSE}
library(ggplot2)
ggplot(trigram_freqs[1:150,], aes(x = 1:150, y=freq)) + geom_bar(stat = "identity")
```

There's a long tail, but the first 30 or so are more frequent than the rest. If
we search for these trigrams across all the errors, we find they actually cover
a substantial portion:

```{r how-many}
tri_regex = paste0("(", paste0(trigram_freqs$trigram[1:30], collapse="|"), ")") %>%
  stri_replace_all_fixed("(", "\\(") %>%
  stri_replace_all_fixed(")", "\\)")
sum(stri_detect_regex(errors, tri_regex))/length(errors)
```

So more than 20% of the Stack Overflow R questions with "Error" contain one of
these first 30 trigrams.  Let's look at some of them:

```{r head}
head(trigram_freqs, 15)
```

If you squint you might understand what these trigrams mean, but a better way to
understand them is to look up a sample of error messages from which they come.

```{r not-find-function}
errors[stri_detect_fixed(errors, "not find function")] %>%
  sample(10) %>% cat(sep="\n---\n")
```

The most frequent error is "could not find function", which is generally due to
either misspelling or failure to load a package properly.

```{r error-in-if}
errors[stri_detect_fixed(errors, "Error in if")] %>%
  head(10) 
```

"Error in if" generally means the logical statement in "`if (XXX) { ...`" is not
yielding a logical value.  Most of these have `missing value where TRUE/FALSE
needed`, meaning that the variable in the `if` statement can't be found.

"could not find" seems to be of the same category as "not find function".

The next four trigrams all seem to come from the same type of error message.

```{r eval-expr}
errors[stri_detect_fixed(errors, "in eval(expr, envir,")] %>%
  sample(10) %>% cat(sep="\n---\n")
```

These are all "object not found errors", where the user has written a statement
that's looking for an object not in memory.

```{r arg-true}
errors[stri_detect_fixed(errors, "= TRUE) :")] %>%
  sample(10) %>% cat(sep="\n---\n")
```
The "`= TRUE) :`" trigram is interesting.  It doesn't seem to match to any
particular error type, but is common because many functions have a logical last
argument.

```{r cannot-open}
errors[stri_detect_fixed(errors, ": cannot open")] %>%
  sample(10) %>% cat(sep="\n---\n")
```
"`: cannot open`" comes from two types of errors: First, a file/connection can't
be opened because R can't find it (likely because of an error in the path), and
second, failures in `.onLoad()` because a package can't find a system dependency.

Let's jump to the next few:

```{r next15}
trigram_freqs[16:30,]
```

New errors in this class are "`object of class`", "`: subscript out`"

```{r object-of-class}
errors[stri_detect_fixed(errors, "object of class")] %>%
  sample(10) %>% cat(sep="\n---\n")
```

These errors all describe the problem of attempting to use a method on an
object for which no such method exists.  It's the object-oriented equivalent
of "can not find function."

```{r subscript-out}
errors[stri_detect_fixed(errors, ": subscript out")] %>%
  sample(10) %>% cat(sep="\n---\n")
```

An easy one: "`subscript out of bounds`". These errors occur when you try to
access an element of a vector or list that isn't there.

```{r failed-for-package}
errors[stri_detect_fixed(errors, "failed for package")] %>%
  sample(10) %>% cat(sep="\n---\n")
```

"`failed for package`" errors are mostly about failure of packages to compile properly,
(often because the user doesn't have the tools installed to build packages from source) Some are due to failure to load due to lack of dependencies.


```{r unable-to-load}
errors[stri_detect_fixed(errors, "unable to load")] %>%
  sample(10) %>% cat(sep="\n---\n")
```

"`unable to load`" errors are also all about missing package dependencies.

```{r is-not-a}
errors[stri_detect_fixed(errors, "is not a")] %>%
  sample(10) %>% cat(sep="\n---\n")
```

"`is not a`" errors are somewhat variable, but most seem to come from cases
where R is expecting an object of one class and recieving another.

```{r err-in-fn}
errors[stri_detect_fixed(errors, "Error in function")] %>%
  sample(10) %>% cat(sep="\n---\n")
```

"`Error in function`" errors are highly variabe in cause. They appear anywhere
an anonymous function fails.

```{r replacement-has}
errors[stri_detect_fixed(errors, ": replacement has")] %>%
  sample(10) %>% cat(sep="\n---\n")
```

"`replacement has`" errors occur when one attempts to assign a vector of values
to a subset of an existing object and the lengths do not match up.

So what have we learned from looking at these errors?  First, a small set of
error types represent an outsize fraction of questions on Stack Overflow.  It
would be prudent to teach the interpretation of these error types in workshops
and classes.

Secondly, almost all of these types of errors have a common theme: *R is looking
for something that isn't there.*  If you look through the errors above, you'll
see R is mostly reporting that it expects to find an object, a logical value, a function, a file,
a sub-object, or a dependency based on the users' input, but that thing isn't there. 
This is an important concept to convey in any "error interpretation" lesson.

*(Comment in the issues [here](https://github.com/noamross/zero-dependency-problems/issues/19)!)*
